Corotating Clumps Around Adolescent Low-Mass Stars: Four Years of Complex Quasiperiodic Variables from TESS

### 2023-08-16 paper/ms.tex
update ms

% primary authors
\author{---AUTHOR LIST \& ORDER NOT FINAL}
\author[0000-0002-0514-5538]{Luke~G.~Bouma}
\altaffiliation{51 Pegasi b Fellow}
\affiliation{\caltech}

@@

The CQV light curves cannot be explained by pure starspots.  Although
all young M dwarfs are spotted, this tends to produce flux variations
over characteristic timescales of the rotation period, $P_{\rm rot}$,
and its half-harmonic, $0.5\,P_{\rm rot}$.  If one fine-tunes the
viewing geometry, starspots can produce dips with durations as short
as $\approx$$0.2\,P_{\rm rot}$, and the dip amplitudes quickly
decrease due to limb-darkening (see \citealt{2017AJ....153..152S},
Figures 37-41).  The observed dips, however, happen over durations as
short as $0.05\,P_{\rm rot}$ with characteristic depths of a few
percent; a ``starspot-only'' scenario can be discarded for any object
with dips that are sufficiently deep and short in duration
\citep{2017AJ....153..152S,2021MNRAS.500.1366K}.   The timescales and
amplitudes of the flux variations in such objects instead require
sharp geometries with material extrinsic to the stellar surface
\citep[e.g.][]{2017AJ....153..152S,2022AJ....163..144G}.  Since the
CQVs are probably all explained by the same phenomenon, we therefore
discard the ``spot-only'' scenario.

@@

To assess the completeness of the resulting 120-second cadence data
that is the basis of this study, we cross-matched TIC8
\citep{2018AJ....156..102S} against the Gaia DR2 point-source catalog
\citep{2018A&A...616A...1G}.  We opted for Gaia DR2 rather than DR3
because the base catalog for TIC8 was Gaia DR2, which facilitated a
one-to-one crossmatch using the Gaia source identifiers.  This
exercise showed that for $T$$<$16 M dwarfs, the TESS 2-minute data are
$\approx$50\% complete at $\approx$50\,pc.  At $<$20\,pc,
$\gtrsim$80\% of the $T$$<$16 M dwarfs have at least one sector of
short-cadence data; at $>$100\,pc, $\lesssim$10\% of such M dwarfs
have at least one sector of short-cadence data.  Armed with this
understanding, we then used our cross-match between Gaia DR2 and TIC8
to select our stars of interest, which we defined as stars with
120-second cadence TESS light curves satisfying these four conditions:
\begin{align}
  T &< 16 \quad&(\mathrm{Bright\ for\ TESS})\\
  \bprp &> 1.5 \quad&(\mathrm{Red\ stars\ only})\\
  M_{\rm G} &> 4 \quad&(\mathrm{Dwarf\ stars\ only})\\
  d &< 150\,{\rm pc} \quad&(\mathrm{Close\ stars\ only}).
\end{align}
Here, $M_{\rm G} = G + 5\log(\varpi_{\rm as}) + 5$ is the Gaia
$G$-band absolute magnitude, $\varpi_{\rm as}$ is the parallax in
units of arcseconds, and $d$ is a geometric distance defined by
inverting the parallax and ignoring any zero-point correction.  The
target sample therefore includes \nstarssearched\ M dwarfs and late-K
dwarfs, down to $T$$<$16 and out to $d$$<$150\,pc.

@@

\begin{figure*}[!tp]
	\begin{center}
		\centering
		\includegraphics[width=0.98\textwidth]{f3.pdf}
    \vspace{-0.3cm}
		\caption{
      {\bf CQVs found in the TESS 2-minute data.}
      Phased TESS light curves over one month are shown for \ngoods\
      CQVs in the high quality sample.  Gray are raw 2-minute data;
      black bins to 300 points per cycle.  Each panel is labeled by
      the TIC identifier, the TESS sector number, and the period in
      hours.  Objects are ordered such that sources with the most TESS
      data available are on top (see Section~\ref{sec:catalog}).
      Sources with large apparent scatter in gray typically show
      secondary periods from unresolved binaries, which are enumerated
      in Table~\ref{tab:thetable}.
		}
		\label{fig:cqvs}
	\end{center}
\end{figure*}

@@

The main types of binarity of interest in this work were those that
could confuse our understanding of CQVs.  For instance, within the
21$''$ TESS pixels, binary companions can produce multiple photometric
signals and hinder our ability to correctly identify the star hosting
the CQV signal.  Binary companions can also bias photometric magnitude
and color measurements, which would affect our stellar parameter
estimates.  To attempt to identify binaries, we considered the
following lines of information.

@@

Of the 27 CQVs in Figure~\ref{fig:evoln}, a few show clear signs of
the CQV phenomenon in one sector, and marginal or non-existent signs
in the other.  While there is subjectivity in this assessment, to our
eyes cases for which at least one sector would be flagged as
``ambiguous'' include
TIC~368129164 (Sector 23 might be labeled an EB),
TIC~177309964 (Sector 38 would be simply a rotating star),
TIC~404144841 (Sector 38 looks like a rotating star),
TIC~201898222 (Sector 3 looks like a rotating star),
TIC~144486786 (Sector 32 might be an RS CVn),
and
TIC~38820496 (Sector 28 might be an RS CVn).
TIC~193831684, assessed on a single-sector basis, would probably be
labeled an eclipsing binary---in fact, \citet{2021ApJ...912..123J}
implicitly have already given this source such a label.  However,
based on the the shape evolution between Sectors 13 and 39, it is a
CQV.  Based on the fraction of sources overall that ``turned off'',
the observed shape evolution implies that CQVs have a duty cycle of
$\approx$75\%.  This type of correction is worth including in
population-level estimates of how intrinsically common CQVs are in the
low-mass stellar population \citep[e.g.][]{2022AJ....163..144G}.

@@

%{\bf AUTHOR CONTRIBUTION STATEMENTS: PLEASE REVISE IF APPROPRIATE)}
LGB conceived the project and performed the dip-counting search, light
curve classification, cluster membership, SED, variability, and
secondary-period analyses.
RJ and SR performed the Fourier-based analysis and contributed
to light curve classification.
LR cross-examined the light curve classification, and contributed an independent SED analysis.
AD identified the magnetic B star connection.
LAH contributed to data analysis and interpretation.
G\'AB acquired and maintained the servers used to run the dip-finding pipeline.
GRR is an architect of the TESS mission.
{\bf JNW?  MNG?}
All authors assisted in manuscript writing and revision.

>>>>

% primary authors
\author[0000-0002-0514-5538]{Luke~G.~Bouma}
\altaffiliation{51 Pegasi b Fellow}
\affiliation{\caltech}

@@

The CQV light curves cannot be explained by pure starspots.  Although
all young M dwarfs are spotted, this tends to produce flux variations
over timescales of the rotation period, $P_{\rm rot}$,
and its half-harmonic, $0.5\,P_{\rm rot}$.  With fine-tuned
viewing geometries, starspots can produce dip durations as short
as $\approx$$0.2\,P_{\rm rot}$, and the dip amplitudes in such
geometries are small
due to limb-darkening (see \citealt{2017AJ....153..152S},
Figures 37-41).  The observed dips occur over durations as
short as $0.05\,P_{\rm rot}$ with characteristic depths of a few
percent; a ``starspot-only'' scenario can be discarded for any object
with dips that are sufficiently large in depth and short in duration
\citep{2017AJ....153..152S,2021MNRAS.500.1366K}.   The dip durations and
amplitudes instead require
sharp geometries with material extrinsic to the stellar surface
\citep[e.g.][]{2017AJ....153..152S,2022AJ....163..144G}.  Since some CQVs require
such geometries, and the
CQVs probably all have a single explanation, we therefore
discard the ``spot-only'' scenario.

@@

To assess the completeness of the resulting 120-second cadence data
that is the basis of this study, we cross-matched TIC8.2
\citep{2019AJ....158..138S,2021arXiv210804778P} against the Gaia DR2 point-source catalog
\citep{2018A&A...616A...1G}.  We opted for Gaia DR2 rather than DR3
because the base catalog for TIC8 was Gaia DR2, which facilitated a
one-to-one crossmatch using the Gaia source identifiers.  This
exercise showed that for $T$$<$16 M dwarfs, the TESS 2-minute data are
$\approx$50\% complete at $\approx$50\,pc.  At $<$20\,pc,
$\gtrsim$80\% of the $T$$<$16 M dwarfs have at least one sector of
short-cadence data; at $>$100\,pc, $\lesssim$10\% of such M dwarfs
have at least one sector of short-cadence data.  Armed with this
understanding, we then used our cross-match between Gaia DR2 and TIC8
to select our stars of interest, which we defined as stars with
120-second cadence TESS light curves satisfying these four conditions:
\begin{align}
  T &< 16 \quad&(\mathrm{Bright\ for\ TESS})\\
  \bprp &> 1.5 \quad&(\mathrm{Red\ stars\ only})\\
  M_{\rm G} &> 4 \quad&(\mathrm{Dwarf\ stars\ only})\\
  d &< 150\,{\rm pc} \quad&(\mathrm{Close\ stars\ only}).
\end{align}
Here, $M_{\rm G} = G + 5\log(\varpi_{\rm as}) + 5$ is the Gaia
$G$-band absolute magnitude, $\varpi_{\rm as}$ is the parallax in
units of arcseconds, and $d$ is a geometric distance defined by
inverting the parallax and ignoring any zero-point correction.  The
target sample therefore includes \nstarssearched\ M dwarfs and late-K
dwarfs, down to $T$$<$16 and out to $d$$<$150\,pc.

@@

\begin{figure*}[!tp]
	\begin{center}
		\centering
		\includegraphics[width=0.98\textwidth]{f3.pdf}
    \vspace{-0.3cm}
		\caption{
      {\bf CQVs found in the TESS 2-minute data.}
      Phased TESS light curves over one month are shown for \ngoods\
      CQVs in the high quality sample.  Gray are raw 2-minute data;
      black bins to 300 points per cycle.  Objects are ordered such
      that sources with the most TESS data available are on top (see
      Section~\ref{sec:catalog}).  
      Each panel is labeled by
      the TIC identifier, the TESS sector number, the period in
      hours, and the three-bit binarity flag from Table~\ref{tab:thetable},
      which denotes Gaia DR3 \texttt{radial\_velocity\_error} outliers
      (bit 1), Gaia DR3 \texttt{ruwe} outliers (bit 2), and stars with
      secondary TESS periods (bit 3). 
		}
		\label{fig:cqvs}
	\end{center}
\end{figure*}

@@

The main types of binaries of interest in this work were those that
were unresolved, because they could confuse our understanding of CQVs.
For instance, unresolved binaries might produce multiple photometric
signals and hinder our ability to correctly identify the star hosting
the CQV signal.  Unresolved binaries could also bias photometric magnitude
and color measurements, which would affect our stellar parameter
estimates.  To attempt to identify binaries, we considered the
following lines of information.

@@

A few objects in Figure~\ref{fig:evoln}
show clear signs of
the CQV phenomenon in one sector, and marginal or non-existent signs
in the other.  While there is subjectivity in this assessment, to our
eyes cases for which at least one sector would be flagged as
``ambiguous'' include
TIC~368129164 (Sector 23 might be labeled an EB),
TIC~177309964 (Sector 38 would be simply a rotating star),
TIC~404144841 (Sector 38 looks like a rotating star),
TIC~201898222 (Sector 3 looks like a rotating star),
TIC~144486786 (Sector 32 might be an RS CVn),
and
TIC~38820496 (Sector 28 might be an RS CVn).
TIC~193831684, assessed on a single-sector basis, would probably be
labeled an eclipsing binary---in fact, \citet{2021ApJ...912..123J}
already gave this source such a label.  However,
based on the the shape evolution between Sectors 13 and 39, it is a
CQV.  Based on the fraction of sources overall that ``turned off'',
the observed shape evolution implies that CQVs have a duty cycle of
$\approx$75\%.  This type of correction is worth including in
population-level estimates of how intrinsically common CQVs are in the
low-mass stellar population \citep[e.g.][]{2022AJ....163..144G}.

@@

%{\bf AUTHOR CONTRIBUTION STATEMENTS: PLEASE REVISE IF APPROPRIATE)}
LGB conceived the project and performed the dip-counting search, light
curve classification, cluster membership, SED, variability, and
secondary-period analyses.
RJ and SR performed the Fourier-based analysis and contributed
to light curve classification.
LR cross-examined the light curve classification, and contributed an independent SED analysis.
AD identified the magnetic B star connection.
LAH contributed to project design and to the interpretation of the light curves.
G\'AB acquired and maintained the servers used to run the dip-finding pipeline.
GRR is an architect of the TESS mission.
{\bf JNW?  MNG?}
All authors assisted in manuscript writing and revision.
